---
title: 'Maize and Beans Yield Response to Fertilizer Types (A and B) in Bugarama Region  '
author: "mailto:jacques.ntezimana@oneacrefund.org"
date: "August 2017"
output:
  html_document:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 6
  word_document:
    toc: yes
    toc_depth: '6'
subtitle: Bugarama RCBD Trial
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries
library(data.table)
library(ggplot2)
library(reshape2)
library(plyr)
library(stats)
library(lattice)
library(gridExtra)
library(pwr)
library(rmarkdown)
library(car)
# define function that turns decimal percentages into pretty formats
format_pct <- function(num) {round(num*100, digits = 2)}

# define function to adjust table widths
html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)
}


```

# Summary 

## Overview 

For profitable production, Beans and Maize require adequate fertilizer application. Applying the right fertilizer at the right rate and at the right time is crucial for a successful crop. The aim of this report is  to determine which type  of fertilizer could lead to the highest Yield  for Maize and/or Beans in the Bugarama region. Maize and Beans Fertilizer application field experiment was carried out on 1000 farmers (i.e 500 farmers for each grade ofe fertilizer).

# Trial Hypothesis 

H1:The type B fertilizer  will produce significantly higher Maize yields than type A fertilizer. 

H2:The type B fertilizer  will produce significantly higher Beans yields than type A fertilizer. 

# Main Take Away

## Summary of Results 

*H1:Did Fertilzer B produce significance high Maize yields than FertilizerA? * Results have shown that Fertilizer B *(mean=773kg/acre, median=762.7kg/acre , 1st.Q-3rd.Q =[564.4-961])* had statistically significant  higher Maize yields than Fertilizer A *(mean=698.6kg/acre, median=700kg/acre , 1st.Q-3rd.Q =[500-900])*.Also  with power calculation , the findings have shown that the sample size used for this study was appropriete at 98% certainly to make correct inference that the sample groups of data are different. 

*H2: Did Fertilizer B produce significance high Beans yields than Fertilizer A ?* Results  revealed that there is no difference in the mean yields of beans  between fertilizer A *(mean=98.9kg/acre , CI=[97.1-101])* and 
B *(mean=101kg/acre , CI=[98.9-102])* . This study may be was underpowered (few participants) to detect a difference of the specified effect size. 

## Data Quality Disclaimer

For the beans yield , we used  statistical outlier detection (Standard Deviation) to detect outliers but we did not consider to eliminate those points from the data . But we did run the analysis with and without it to se how the results change. 


# Recommadation 

.	We therefore conclude that we should recommend to farmers that they adopt Fertilizer B  over Fertilizer A for Maize. 

.	For Beans crop , fertilizer B  had yields which were much higher than the Fertilizer A yield, but the results were not statistically significant (due to low sample size); this is the reason why no types of fertilizer is highlighted for that crop.Also , it doesn't make sense to use the same sample size for both Beans and Maize to detect the same effect since they do not have the same capacity of nutrient fixation . 

.	We ought to re-do  Beans fertilizer trial but consider to to increase the sample size . Given this risk, it is not acceptable to conclude that fertilizing beans has no impact when the sample size used for the research was not sufficient to detect a meaningful difference between the fertilizer type A and B.

.In meanwhile ,we also recommend the researcher to run ex-ante power calculation to  determine the minimum sample size required to meet power level and minimum detectable effect set for the study.

# Analysis 

```{r assignment}
# create directory 

wd <- "C:/Users/Admin/Desktop/r_directory/AInnovations/Analyst Mentorship Program/AMP"
dd <- paste(wd, "data", sep="/")
od <- paste(wd, "output", sep="/")

## call data 
c <- read.csv(paste(dd, "Homework_data.csv", sep = "/"), header = TRUE, stringsAsFactors = FALSE) ## cases 

```

## Diagonostic testing 

### Data exploration by box-plots

```{r boxplots}
# data  visualization for outlier check

g1 <- ggplot(c, aes(x=Group, y=Maize.yield.per.acre, fill=Group)) + geom_boxplot() +
    guides(fill=FALSE) + ylab("Maize yields kg per acre")+ggtitle("Maize Fertilizer types application")

g2 <- ggplot(c, aes(x=Group, y=Beans.yield.per.acre, fill=Group)) + geom_boxplot() +
    guides(fill=FALSE) + ylab("Beans  yields kg per acre")+ggtitle("Beans Fertilizer types application")

grid.arrange(g1, g2, ncol=2) 

```

Typically, box-plot allow us to see the distribution of a variable data and whether they contain outliers or not.  Indeed, those plotted box-lots  are helping us see the performance of Fert.A and Fert.B on Maize and Beans Yields(kg) per Acre.Results revealed that Fert.B has good performance compared to Fert.A for both Maize and Beans yields. But in this particular case, box plots have helped to investigate something specific. It reveals some information on the outliers for beans yields that there're large numbers  of kg  that max out near 160 for both fertilizer types and also has some outlier points at the low end . 

Since some of the Beans data are unusual compared to the others which can have strong influence on the results, we should try to identify the cause of the unusual nature points. we can however choose to correct any data entry or measurement error, then consider removing data that are associated with special causes and repeating analysis.

### Test of normality of data 
#### QQ-plot

```{r QQ-plot}
# QQ plot  all data ( both group togetheer )

par(mfrow = c(1, 2))
qqnorm(c$Maize.yield.per.acre, main="QQ plot Maize Yield ")
qqline(c$Maize.yield.per.acre)

qqnorm(c$Beans.yield.per.acre, main="QQ plot Beans Yield")
qqline(c$Beans.yield.per.acre)


## QQ-plots for subgroup 
## create subsets

c1 <- subset(c,Group=="Fert.A", select=ID:Group)
c2 <- subset(c,Group=="Fert.B", select=ID:Group)

### maize
par(mfrow = c(1, 2))
qqnorm(c1$Maize.yield.per.acre, main="QQ plot Maize Yield  Fert.A")
qqline(c1$Maize.yield.per.acre)

qqnorm(c2$Maize.yield.per.acre, main="QQ plot Maize Yield  Fert.B")
qqline(c2$Maize.yield.per.acre)

### Beans
par(mfrow = c(1, 2))
qqnorm(c1$Beans.yield.per.acre, main="QQ plot Beans Yield  Fert.A")
qqline(c1$Beans.yield.per.acre)

qqnorm(c2$Beans.yield.per.acre, main="QQ plot Beans Yield  Fert.B")
qqline(c2$Beans.yield.per.acre)


```

As QQ-plot plots a perfect normal distribution as a solid diagonal line and then our actual data as markers (circles) on top of that. Therefore, the plotted  entire data(Fert A&B) for beans yield shows some normal distribution while Maize yield data shows non-normal distribution trend. 

However the plots look well when data are plotted in subgroups. The QQ-plots for Beans show that both Fert.A and Fert.B data follow a normal distribution while examining the data in plot of Maize yield,Fertilizer A has some larger deviation(its actual points deviate from the normal line ) than Fert.B . 

How do we be sure with this discrepancy? Le's run Shapiro test to see whether those systematic deviations from the normal line exist or not .

#### Shapiro-Wilk normality test

```{r test of normality}

## Shapiro test  
## Just to remember that c is  a dataframe for all groups data, c1 is subset for Fert.A only ,c2 IS for Fer.B 

shapiro.test(c$Maize.yield.per.acre) 
shapiro.test(c$Beans.yield.per.acre)

shapiro.test(c1$Maize.yield.per.acre) 
shapiro.test(c2$Maize.yield.per.acre) 

shapiro.test(c1$Beans.yield.per.acre)
shapiro.test(c2$Beans.yield.per.acre)

```

Our Maize yields distribution has a Shapiro test P-value <5%  for all  Fertilizer types A&B tested separately and when tested both together. This is strongly suggests  that our Maize data is non- normal. While  Beans yields data,  Shapiro test P-Values > 5%  this indicates that data are normal. 

## Test of Significance 

Since the results from Shapiro test suggests the normality for Beans Data and non-normal for Maize data, it is recommended to use different test statistic.  Non-parametric test is recommended for non-normal data, vs parametric test when data are normal. Therefore ,to test the mean difference between Fert.A  and fert B for Beans data which are  normal , Welch Test( An improvement on the T-test, which is more robust to unequal variances)  is recommended while the Wilcoxon Test is recommended for Maize Yield data that  are not normal.

### Welch Test for Beans Yield Data


```{r test of mean difference }

## Welch Test 
t.test(Beans.yield.per.acre~Group, alternative = "two.sided", var.equal = FALSE, data=c)$p.value #perform the Welch test and store p-value

## Let us verify results using T test 

t.test(Beans.yield.per.acre~Group, alternative = "two.sided", var.equal = TRUE, data=c)$p.value # We have gotten similar result because the variance is almost the same
 
```

No siginficance difference mean yield found  between two types of fertilizer applied to Beans ( Welch Test p-values>5%)

### Wilcoxon Test  for  Maize Yield Data

```{r test of mean difference wilcoxon }

## Wilcoxon Test 

wilcox.test(Maize.yield.per.acre~Group,paired=FALSE, correct=TRUE,data=c)

## let us verify results using T.test
t.test(Maize.yield.per.acre~Group, alternative = "two.sided", var.equal = TRUE, data=c) # has produce significance which should not be always promises , we must  use the right test at the right time. 

```

We found a statistically  significance  mean yield difference between two types of fertilizer applied to Maize ( Wilcoxon Test p-values<5%)

## Power Analysis  

Our results have shown no difference in mean yield for Beans among  Fertilizer A and B ( Welch test P-values=0.165 > 5%)  and found to have statistically significant difference for Maize (Wilcoxon Test p-value = 4.565e-05 <5%) . However , there is a need of performing  power calculations to assess whether the presence  or lack of impact difference is due to the sample size, inherent to the intervention itself, or due to some other reason such as implementation failure or something else. 

### Power calculation 

It is important that we use the correct power calculation in much the same way it is important we use the right hypothesis test.A low power will mean our results are less reliable and we may be making false conclusions about interventions.The required sample size to detect the effect of interest  will need to produce the power of 0.8 and above . Power calculations are depending on the data distribution .Therefore  with normal data ,* pwr.2p2n.test* will be used while *Mcpower* will be used for Non normal data . 

Also since sample size, power and effect size are all part of the same equation and all related, We will first need to calculate effect size which is difference in means between samples divided by the standard deviation of the samples  using *Cohen's D equation* .

#### Power Calculation  for normal ( Beans Yield )

```{r normal}

# Generate effect size using Cohen D equation  (Note that c1 is dataframe that contain Data for Fert.A while c2 contain data for Fert.B)
#Don't forget to re run the command that generate c1 and c2 in above codes

cohen_d <- function(d1,d2) {  
  m1 <- mean(d1, na.rm=TRUE)
  m2 <- mean(d2, na.rm=TRUE)
  s1 <- sd(d1, na.rm=TRUE)
  s2 <- sd(d2, na.rm=TRUE)
  spo <- sqrt((s1**2 + s2**2)/2)
  d <- (m1 - m2)/spo
  effsi <- d / sqrt((d**2)+4)
  ret <- list("d" = d, "effectsi" = effsi)
  return(ret)  } 

##get effect size and call it like this:
 
vals<- cohen_d(c1$Beans.yield.per.acre, c2$Beans.yield.per.acre)
 
pwr.2p2n.test(h = vals$effectsi , n1 =length(c1$Group), n2=length(c2$Group) , sig.level = 0.05, power = NULL)

```

With the result calculated above , a statistical power of 0.106 was obtained , which means a weak power . Therefore ,it can be said with only 11% of certainly that there is no really a statistically significant difference between the beans'mean yields for two types of fertilizers . so we we can conclude that the sample size used was inappropriate for detecting whether meaningful effects have occurred or not. Given this risk, it is not acceptable to conclude that an intervention has not impact when the sample size used for the research is not sufficient to detect a meaningful difference between two types of fertilizer.

#### Power calculation  for Non-normal ( Maize Yield )

```{r nonnormal}

MCpower = function(sample1, sample2, size) {
  reps = 1000
    results  <- sapply(1:reps, function(r) {
        resample1 <- sample(sample1, size=size, replace=TRUE) 
        resample2 <- sample(sample2, size=size, replace=TRUE) 
        test <- wilcox.test(resample1, resample2, alternative="two.sided",paired=FALSE, correct=TRUE)
        test$p.value
    })
    sum(results<0.05)/reps
}

 MCpower(c1$Maize.yield.per.acre, c2$Maize.yield.per.acre, 500)

```

With the result calculated above , a statistical power of 0.979 was obtained , which means a high power . Therefore ,it can be said that we are confident at 98% certainly that there is a statistically significant difference between the Maize'mean yields between two types of fertilizers . 

# Follow up with Outlier for Beans Data 

## Run Outlier Check, QQ-plots and shapiro Test 

Since our beans data has shown the outlier data points data, it is crucial to run again results to see if the outliers had any influential on the results .

```{r outlier}

## run outliers check function and detect it 

outlier.check <- function(x) {
  
  name <- ifelse(x < (mean(x, na.rm=T)-(2*sd(x, na.rm=T))) | 
                   x > (mean(x, na.rm=T) + (2*sd(x, na.rm=T))), 1, 0)
  
  return(name)
}

c$Beans_yield.check <- outlier.check(c$Beans.yield.per.acre)

## check again normality by considering outliers removal 

##subset data without outliers 

c0 <- subset(c,Beans_yield.check==0, select=ID:Beans_yield.check) ## all database without outliers 
c3 <- subset(c0,Group=="Fert.A", select=ID:Beans_yield.check)     ## database for Fert A  without outliers  
c4 <- subset(c0,Group=="Fert.B", select=ID:Beans_yield.check)     ## database for Fert B  without outliers
##
### qq-plots beans data with out outliers
par(mfrow = c(1,3))
qqnorm(c0$Beans.yield.per.acre, main="QQ plot Beans Yield  ")
qqline(c0$Beans.yield.per.acre)

qqnorm(c3$Beans.yield.per.acre, main="QQ plot Beans Yield  Fert.A")
qqline(c3$Beans.yield.per.acre)

qqnorm(c4$Beans.yield.per.acre, main="QQ plot Beans Yield  Fert.B")
qqline(c4$Beans.yield.per.acre)

## Run Shapiro test for data with out outliers 

shapiro.test(c0$Beans.yield.per.acre)
shapiro.test(c3$Beans.yield.per.acre)
shapiro.test(c4$Beans.yield.per.acre)


```

Huu, very interesting !!!. Those are some terrible when you look only on your data without taking into the oultiers . Some reality  now start to happening .Before ,  shapiro test had shown us that data is normal , then the results have shown non -normal trend after removing outliers *shapiro test p-values <5%* which suggests us to use non parametric test .

## Run Wilcoxon Test and Power Calculation 

```{r nonnormal2}

## wicoxon test 

wilcox.test(Beans.yield.per.acre~Group,paired=FALSE, correct=TRUE,data=c0)

## MCpower  

MCpower = function(sample1, sample2, size) {
  reps = 1000
    results  <- sapply(1:reps, function(r) {
        resample1 <- sample(sample1, size=size, replace=TRUE) 
        resample2 <- sample(sample2, size=size, replace=TRUE) 
        test <- wilcox.test(resample1, resample2, alternative="two.sided",paired=FALSE, correct=TRUE)
        test$p.value
    })
    sum(results<0.05)/reps
}

 MCpower(c3$Beans.yield.per.acre, c4$Beans.yield.per.acre, 500)  ## number of database are not equal 

```

Now , we can conclude that there is no significance difference but sample size used for the research was not sufficient to detect a meaningful difference between two types of fertilizer. 

# Homework

*With Beans data , What is an optimal sample size required to detect minimum detectable effect (MDE) of interest since we know that having a sample size smaller than statistically required increases the likelihood of researchers concluding that the evaluated intervention has no impact when the intervention does  Also for cost efficiency and value-for-money it is important to ensure that an evaluation research design does not use a larger sample size than is required to detect the minimum detectable effect (MDE) of interest*

# Summary Stats  for Reporting 

```{r summary }

## for normal data , report means and confidence Interval  
library(rcompanion)

groupwiseMean(Beans.yield.per.acre ~ Group, 
              data   = c, 
              conf   = 0.95, 
              digits = 3)

## for Non normal data ,report Median and ( 1st and 3rd quartiles) 

setnames(c, old = c('Maize.yield.per.acre'),new = c('MaizeYield')) ## keep variable name short

aggregate(MaizeYield ~ Group, data = c, summary)

```
